---
title: "Impact of Severe Weather on Communities and Municipalities"
output: html_document
---

## Synopsis

Bacon ipsum dolor sit amet pig pancetta ground round short loin tenderloin cow
biltong leberkas chicken kielbasa beef ribs porchetta swine meatloaf. Ball tip
short loin sausage porchetta, tongue capicola ham beef. Cow chuck prosciutto,
frankfurter pork pastrami biltong leberkas chicken meatball bacon. Jowl chuck
fatback tail meatball tongue ham hock capicola prosciutto kevin brisket pork
chop. Spare ribs frankfurter pancetta ribeye, tri-tip ham sausage pig. T-bone
chuck landjaeger cow ground round shankle filet mignon turducken drumstick
capicola spare ribs.  

## Data Processing

The [Storm Data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2)
file is downloaded if a local copy does not already exists. The `read.csv` function
can directly read the `.csv` file embedded in the downloaded file.

```{r datain, echo = TRUE, cache = TRUE}

# download file (if it is necessary)
fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
localFilename <- "StormData.bz2"
if (!file.exists(localFilename)) {
    download.file(fileUrl, destfile = localFilename, mode = "wb")
} 

# read the csv file embedded in the .bz2 file
StormData <- read.csv(localFilename)

# size of the imported data set
dim(StormData)

# assign formats to specific columns
StormData$BGN_DATE <- as.Date(StormData$BGN_DATE, format = "%m/%d/%Y" )
StormData$EVTYPE <- as.factor(StormData$EVTYPE)


```

First, those variables that have no recorded injuries nor fatalities plus
no crop or property damage are identified. This permits dropping slightly more
than 70% of the data for this analyis.  

Even in this subset of data, the fields are completed in a very inconsistent
manner. It is the opinion of this analysis that refresher training be provided
to those recording the data to encourage the data recorded in a more uniform
manner. The data will become more trustworthy and require less conjecture to
prepare analytic data set suitable for study.  

Property and Crop damage entries with `+` or `-` in the exponent field have
been imputed with an exponent of 0. Investigating the few cases where this
occurs, the values for property and crop damage are probablly grossly
underestimated. However as two fatalities occur in the single variable with
`-` for the exponent, I prefer not to drop the variables.  

```{r tidyup, echo=TRUE, cache=FALSE}
# tidy up parts off the data

# drop variables that have no effect (no damage, no injuries, no fatalities)
zeroPropDmg <- StormData$PROPDMG %in% 0
zeroCropDmg <- StormData$CROPDMG %in% 0
zeroFatal <- StormData$FATALITIES %in% 0
zeroInjury <- StormData$INJURIES %in% 0

zeroEffect <- (zeroPropDmg & zeroCropDmg & zeroFatal & zeroInjury)

StormData <- StormData[!zeroEffect,]

# size of the data set after dropping nonfunctional data

dim(StormData)

# now cleanup the exponents

# we have these chars to cleanup in PROPDMGEXP:
#
# [1] "K" "M" ""  "B" "m" "+" "0" "5" "6" "4" "h" "2" "7" "3" "H" "-"

# we have these chars to cleanup in CROPDMGEXP:
#
# [1] ""  "M" "K" "m" "B" "?" "0" "k"

print(paste(sum(StormData$PROPDMGEXP %in% "+"), " variable with + as exponent for Property Damage"))
print(paste(sum(StormData$PROPDMGEXP %in% "-"), " variable with - as exponent for Property Damage"))

expSubs <- list(
    c("K",3),
    c("M",6),
    c("^$",0),
    c("m",6),
    c("\\?",0),
    c("h",2),
    c("H",2),
    c("B",9),
    c("k",3),
    c("[+]",0),
    c("[-]",0))

StormData$NewPROPDMGEXP <- StormData$PROPDMGEXP
StormData$NewCROPDMGEXP <- StormData$CROPDMGEXP

for (i in expSubs) {
    StormData$NewPROPDMGEXP <- gsub(i[1],i[2],StormData$NewPROPDMGEXP)
    StormData$NewCROPDMGEXP <- gsub(i[1],i[2],StormData$NewCROPDMGEXP)
}

StormData$NewPROPDMGEXP <- as.numeric(StormData$NewPROPDMGEXP)
StormData$NewCROPDMGEXP <- as.numeric(StormData$NewCROPDMGEXP)

dim(StormData)
str(StormData)

library(plyr)
summ1 <- ddply(StormData, "EVTYPE", summarize,
               TotPropertyDamage = sum(PROPDMG * 10^NewPROPDMGEXP),
               TotCropDamage = sum(CROPDMG * 10^NewCROPDMGEXP),
               TotFatalities = sum(FATALITIES),
               TotInjuries = sum(INJURIES))
summ1$sumPropCropDamage <- summ1$TotPropertyDamage + summ1$TotCropDamage
summ1$TotPropertyDamage <- summ1$TotPropertyDamage / 1e9
summ1$TotCropDamage <- summ1$TotCropDamage / 1e9
summ1$sumPropCropDamage <- summ1$sumPropCropDamage / 1e9
head(summ1)

```

## Results

### Cleveland Plots of Fatalities + Injuries vs EVTYPE:  

```{r foobar, echo = TRUE}

dim(summ1)
# zeroFatalities <- summ1$TotFatalities == 0 | is.na(summ1$TotFatalities)
# zeroInjuries <- summ1$TotInjuries == 0 | is.na(summ1$TotInjuries)
# 
# sum(zeroFatalities)
# sum(!zeroFatalities)
# sum(zeroInjuries)
# sum(!zeroInjuries)
# 
# unique(summ1$TotFatalities)
# unique(summ1$TotInjuries)
# 
topFatalities <- order(summ1$TotFatalities, decreasing = TRUE, na.last = TRUE)
topInjuries <- order(summ1$TotInjuries, decreasing = TRUE, na.last=TRUE)
topPropDamage <- order(summ1$TotPropertyDamage, decreasing = TRUE, na.last=TRUE)
topCropDamage <- order(summ1$TotCropDamage, decreasing = TRUE, na.last=TRUE)

require(ggplot2)
require(scales)

summ1$TotFatalities[order(summ1$TotFatalities, decreasing = TRUE, na.last = TRUE)[1:25]]
topNN <-  25
p1 <- ggplot(summ1[topFatalities[1:topNN],],
             aes(x=TotFatalities, y=reorder(EVTYPE, TotFatalities))) +
    geom_segment(aes(yend=reorder(EVTYPE,TotFatalities)), xend=0, colour="grey50") +
    geom_point(size=3) +
    theme_bw() +
    theme(panel.grid.major.y = element_blank()) +
    geom_text(aes(label=TotFatalities), hjust=-0.5, colour="red")
p1 <-p1 + xlab("Total Number of Fatalities") +
    ylab("Event Type")
p1 + scale_x_log10(limits=c(1,80000))

summ1$TotInjuries[order(summ1$TotInjuries, decreasing = TRUE, na.last=TRUE)[1:25]]
p2 <- ggplot(summ1[topInjuries[1:topNN],],
             aes(x=TotInjuries, y=reorder(EVTYPE, TotInjuries))) +
    geom_segment(aes(yend=reorder(EVTYPE,TotInjuries)), xend=0, colour="grey50") +
    geom_point(size=3) +
    theme_bw() +
    theme(panel.grid.major.y = element_blank()) +
    geom_text(aes(label=TotInjuries), hjust=-0.5, colour="red")
p2 <- p2 + xlab("Total Number of Injuries") +
    ylab("Event Type")
p2 + scale_x_log10(limits=c(1,800000))

topCPDamage <- order(summ1$sumPropCropDamage, decreasing = TRUE)
p5 <- ggplot(summ1[topCPDamage[1:topNN],],
             aes(x=sumPropCropDamage, y=reorder(EVTYPE, sumPropCropDamage))) +
    geom_segment(aes(yend=reorder(EVTYPE,sumPropCropDamage)), xend=0, colour="blue") +
    geom_point(size=3, color="blue") +
    geom_text(aes(label=format(sumPropCropDamage, digits=2), hjust=-0.5))

p5 <- p5 + theme_bw() + theme(panel.grid.major.y = element_blank()) +
    xlab("Sum of Total Crop and Property Damage ($Billions)") +
    ylab("Event Type")

p5 + scale_x_log10(limits=c(1,400)) 
    
```